Best Practices for Developing Monte Carlo Methodologies in Molecular Simulations [Article v\versionnumber]

### 2025-06-05 paper/paper.tex
quat->axis_angle

@@

In the example shown in Code Block~\ref{sim:statistics}, the standard deviation is computed using the blocking method \cite{flyvbjerg_error_1989}.
In this case, only blocks of $100$ samples are considered, but a careful analysis would show that the resulting standard deviation is independent of the block size (when large enough).
Thus, with this implementation, we are assuming that the sampled states are independent after $100$ MC trials, which may be reasonable for an ideal gas but most likely not for a dense liquid.
See the blocking tutorial in math plugin of FEASST \cite{hatch_monte_2024} for an example with computing the block standard deviation from the average position of a particle randomly walking in one dimension with periodic boundary conditions.
If the maximum trial displacement is smaller than the periodic boundary, correlations in the position may persist over many trials.

@@

In the example shown in Code Block~\ref{sim:unit_sphere}, positions are generated uniformly on the surface of a sphere.
This is an important case to consider, because MC trials often assume that particles may be randomly oriented.
However, one must be careful in how an orientation is generated.
Uniformly selected values of Euler angles, for example, will result in a bias along the poles.
Code Block~\ref{sim:quaternion} also shows how quaternions may be randomly generated to represent a uniformly random orientation of a point on the surface of a sphere \cite{vesely_angular_1982}.
Code Block~\ref{sim:rotationmatrix} uses these randomly generated quaternions to compute rotation matrices.
A randomly generated perturbation of orientation is then shown in Fig.~\ref{fig:orientation_perturbation}.

@@

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:rotationmatrix},
caption=The rotation\_matrix module generates a rotation matrix from a quaternion.
]{../codes/rotation_matrix.py}
\end{figure}

@@

\begin{figure}
\begin{centering}
\includegraphics[width=3cm]{../figures/random_orientation.png}
\caption{
Example output from Code Block~\ref{sim:random_orientation} shows $10^3$ uniformly random points on a sphere in blue, and $10^3$ points in orange randomly generated about the upwards $z$-axis using an orientational tunable parameter, $\delta=0.3$.
\label{fig:orientation_perturbation}
}
\end{centering}
\end{figure}

@@

\begin{figure}
\begin{centering}
\includegraphics[width=3cm]{../figures/spherical_shell.png}
\caption{
Example output from Code Block~\ref{sim:spherical_shell} shows $10^5$ small blue points randomly generated inside a spherical shell of outer radius $3$ and inner radius $2$.
\label{fig:spherical_shell}
}
\end{centering}
\end{figure}

>>>>

$10^6$ trials, $10^2$ bins, $10$ simulations, $\delta \theta_{max}=\pi/10$.

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:orientation_trials},
caption={Caption.}
]{../codes/orientation_trials.py}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[width=8.5cm]{../figures/orientation_trials.pdf}
\caption{
\label{fig:orientation_trials}
}
\end{centering}
\end{figure}

@@

In the example shown in Code Block~\ref{sim:statistics}, the standard deviation is computed using the blocking method \cite{flyvbjerg_error_1989}.
In this case, only blocks of $100$ samples are considered, but a careful analysis would show that the resulting standard deviation is independent of the block size (when large enough).
Thus, with this implementation, we are assuming that the sampled states are independent after $100$ MC trials, which may be reasonable for an ideal gas but most likely not for a dense liquid.
See the blocking tutorial in python interface of FEASST \cite{hatch_monte_2024} for an example with computing the block standard deviation from the average position of a particle randomly walking in one dimension with periodic boundary conditions.
If the maximum trial displacement is smaller than the periodic boundary, correlations in the position may persist over many trials.

@@

In the example shown in Code Block~\ref{sim:unit_sphere}, positions are generated uniformly on the surface of a sphere.
This is an important case to consider, because MC trials often assume that particles may be randomly oriented.
However, one must be careful in how an orientation is generated.
Uniformly selected values of Euler angles, for example, will result in a bias along the poles.
Code Block~\ref{sim:quaternion} also shows how quaternions may be randomly generated to represent a uniformly random point on the surface of a sphere \cite{vesely_angular_1982}.
Code Block~\ref{sim:rotationmatrix} uses these randomly generated quaternions to compute rotation matrices.
A randomly generated perturbation of orientation is then demonstrated in Fig.~\ref{fig:random_orientation}.

@@

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:rotationmatrix},
caption=The rotation\_matrix module generates a rotation matrix from an axis and angle.
]{../codes/rotation_matrix.py}
\end{figure}

@@

\begin{figure}
\begin{centering}
\includegraphics[width=6cm]{../figures/random_orientation.png}
\caption{
  Example output from Code Block~\ref{sim:random_orientation} shows $10^3$ uniformly random unit vectors emanating from the origin and $10^3$ unit vectors in orange randomly generated about the upwards $z$-axis (shown in black) with axis-angle rotations of a maximum angle parameter, $\delta \theta_{max}=\pi/2=90$ degrees.
\label{fig:random_orientation}
}
\end{centering}
\end{figure}

@@

\begin{figure}
\begin{centering}
\includegraphics[width=8.5cm]{../figures/spherical_shell.png}
\caption{
Example output from Code Block~\ref{sim:spherical_shell} shows approximately $10^5$ small blue points randomly generated inside a spherical shell of outer radius $3$ and inner radius $2$ and cut in half at $z=0$.
\label{fig:spherical_shell}
}
\end{centering}
\end{figure}
